---
title: "RNA_seq_DESeq2"
author: "hputnam"
edited by: "hputnam"
date: "9/02/2021"
output: html_document
---


## RNAseq Differential Expression Analysis 

Load libraries
```{r, message=FALSE, warning=FALSE}

library("tidyverse")
library("genefilter")
library("DESeq2")
library("goseq")

# library("factoextra")
# library("MuMIn")
# library("NbClust")
# library("ComplexHeatmap")
# library("tidyverse")
# library("tidyr")
# library("lme4")
# library("RColorBrewer")
# library("magrittr")
# library("ggplot2")
# library("goseq")
library("pheatmap")
# library("stats")
# library("gridExtra")
# library("VennDiagram")
# library("here")
# library("plotrix")
```

Import the data files 
```{r}
#treatment information
treatmentinfo <- read.csv("RAnalysis/Data/Mcap_RNAseq_metadata.csv", header = TRUE, sep = ",")
rownames(treatmentinfo) <- treatmentinfo$sample_num
str(treatmentinfo)
head(treatmentinfo)

#gene count matrix
gcount <- as.data.frame(read.csv("RAnalysis/Data/Mcap_TRIMMOMATIC_HISAT2_gene_count_matrix.csv", row.names="gene_id"))
dim(gcount)
head(gcount)
colSums(gcount)

# Make sure treatment info ='s the rows of count data
all(rownames(treatmentinfo) %in% colnames(gcount)) # must come out TRUE

```


##### Pre-filter gene counts
## Pre-filtering our dataset to reduce the memory size dataframe, increase the speed of the transformation and testing functions, and improve quality of statistical analysis by removing low-coverage counts. Removed counts could represent outliers in the data and removing these improves sensitivity of statistical tests. We will filter out low coverage samples. 

```{r}

#Do we want to do any filtering?

#why do we want to filter?

#what do we want to filter?

#how do we want to filter

#Prefiltering versus DESeq2 default filtering
#http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#indfilt


filt <- filterfun(pOverA(0.5,10))

#create filter for the counts data
gfilt <- genefilter(gcount, filt)

#identify genes to keep by count filter
gkeep <- gcount[gfilt,]

#identify gene lists
gn.keep <- rownames(gkeep)

#gene count data filtered in PoverA, P percent of the samples have counts over A
gcount_filt <- as.data.frame(gcount[which(rownames(gcount) %in% gn.keep),])
head(gcount_filt)
dim(gcount_filt)


```

#Create a DESeqDataSet design from gene count matrix and labels. Here we set the design to look at treatment to test for any differences in gene expression across timepoints attributed to treatment.
#not using DESeq due to random factors, but can still use this to visualize clusters
```{r}
#Set DESeq2 design
gdds <- DESeqDataSetFromMatrix(countData = gcount_filt,
                              colData = treatmentinfo,
                              design = ~method)
```

#### Visualize gene count data

We're looking to see if the samples of the same treatments cluster

##### Log-transform the count data
First we are going to log-transform the data using a variance stabilizing transforamtion (vst). This is only for visualization purposes. Essentially, this is roughly similar to putting the data on the log2 scale. It will deal with the sampling variability of low counts by calculating within-group variability (if blind=FALSE). Importantly, it does not use the design to remove variation in the data, and so can be used to examine if there may be any variability do to technical factors such as extraction batch effects.

To do this we first need to calculate the size factors of our samples. This is a rough estimate of how many reads each sample contains compared to the others. In order to use VST (the faster log2 transforming process) to log-transform our data, the size factors need to be less than 4. Otherwise, there could be artifacts in our results.
```{r}
SF.gdds <- estimateSizeFactors(gdds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than four to use vst
print(sizeFactors(SF.gdds)) #View size factors
```

Our size factors are all less than 4, so we can use VST!
```{r}
gvst <- vst(gdds, blind=FALSE) #apply a variance stabilizing transformation to minimize effects of small counts and normalize library size
```

#### Principal component plot of samples
```{r}
gPCAdata <- plotPCA(gvst, intgroup = c("method"), returnData=TRUE)
percentVar <- round(100*attr(gPCAdata, "percentVar")) #plot PCA of samples with all data
ggplot(gPCAdata, aes(PC1, PC2, color=method)) + 
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  #scale_color_manual(values = c(PolyA="slateblue", RiboFree="indianred3")) +
  coord_fixed() +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     #panel.grid.major = element_blank(), #Set major gridlines
                     #panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()) + #Set the plot background
  theme(legend.position = ("none")) #set title attributes
```

##### Run DE analysis for overall example 

#Run differential expression test using a Wald model. 
```{r, message = FALSE}
DEGSeq2_wald <- DESeq(gdds, test = "Wald") #run differential expression test by group using the Wald model
```

#Explore significant p-values for treatment
```{r, message = FALSE}
DEGSeq2.results <- results(DEGSeq2_wald, contrast=c("method","PolyA","RiboFree")) 
head(DEGSeq2.results)
sum(DEGSeq2.results$padj < 0.05, na.rm=TRUE) #How many adjusted p-values were less than 0.05

#create data frame with only DESeq2 DEG's with < 0.05 significance 

DEGSeq2.sig.results <- subset(DEGSeq2.results, padj < 0.05)
DEGSeq2.sig.results <- as.data.frame(DEGSeq2.sig.results)
DEGSeq2.sig.results$gene_id <-rownames(DEGSeq2.sig.results)
write.csv(DEGSeq2.sig.results, 'RAnalysis/Output/DEGsig.results.csv')  

```

## Visualize differentially expressed genes from DESeq2
```{r}
# Subset list of genes by those which padj>0.05
DEGSeq2.output <- as.data.frame(DEGSeq2.sig.results)
dim(DEGSeq2.output)
DEGSeq2.results.all <- DEGSeq2.sig.results
DEGSeq2 <- rownames(DEGSeq2.results.all) # list all gene names 
DEGSeq2 <- unique(DEGSeq2) # select only unique gene names 
DEGSeq2_list <- gdds[which(rownames(gdds) %in% DEGSeq2)] # filter gdds DESeq2 object by unique gene names
dim(DEGSeq2_list) # 15 x 15
print(counts(DEGSeq2_list))

# As determined above, size factors all less than 4, so proceed with VST
# apply a variance stabilizing transformation to minimize effects of small counts and normalize library size

DEGvst <- vst(DEGSeq2_list, blind = FALSE, nsub = nrow(counts(DEGSeq2_list)))

dim(DEGvst) # 15 x 15
print(assay(DEGvst)) # look at vst-transformed gene count data 

# Plot heat map with diff expressed genes
# Make a matrix for computing similarity
mat <- assay(DEGvst)#[DEG.results.all, ] # make an expression object
mat <- mat - rowMeans(mat) #difference in expression compared to average across all samples
dim(mat)
ann_colors <- list(treatment= c(PolyA="black", RiboFree="cadetblue3"))
df_DEGSeq2 <- as.data.frame(colData(DEGvst)[c("method")]) #make dataframe for column naming and associated treatment
DEGSeq2_heatmap <- pheatmap(mat, scale= "row", legend=TRUE, annotation_legend=TRUE, annotation_col=df_DEGSeq2, annotation_colors = ann_colors,
                            clustering_distance_rows="euclidean", clustering_method = "average",
                            show_rownames =FALSE,
                            show_colnames =TRUE,
                            cluster_cols = TRUE)

pdf(file = "RAnalysis/Output/DEGSeq2_heatmap.pdf")
DEGSeq2_heatmap
dev.off()


# PCA plot of diff-expressed genes 
DEGSeq2_PCAdata <- plotPCA(DEGvst, intgroup = c("method"), returnData=TRUE)
percentVar_pca <- round(100*attr(DEGSeq2_PCAdata, "percentVar")) #plot PCA of samples with all data
DEGSeq2_PCA_plot <- ggplot(DEGSeq2_PCAdata, aes(PC1, PC2, color=method)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar_pca[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_pca[2],"% variance")) +
  scale_color_manual(values = c(PolyA="black", RiboFree="cadetblue3")) +
  coord_fixed() +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
        #panel.grid.major = element_blank(), #Set major gridlines
        #panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()) #Set the plot background

DEGSeq2_PCA_plot

ggsave(path = "RAnalysis/Output/", filename = "DEGSeq2_PCA_plot.pdf")

# PCA plot is of deferentially expressed genes only

# Save results

write.csv(counts(DEGSeq2_list), file="RAnalysis/Output/DEGSeq2_list_unique.csv")

```



