---
title: "RNA_seq_DESeq2"
author: "hputnam"
edited by: "laurenzane"
date: "5/16/2022"
output: html_document
---


## RNAseq Differential Expression Analysis 

Load libraries
```{r, message=FALSE, warning=FALSE}

library("tidyverse")
library("genefilter")
library("DESeq2")
library("goseq")


# library("factoextra")
# library("MuMIn")
# library("NbClust")
# library("ComplexHeatmap")
# library("tidyr")
# library("lme4")
# library("RColorBrewer")
# library("magrittr")
library("ggplot2")
library("pheatmap")
# library("stats")
# library("gridExtra")
# library("VennDiagram")
# library("here")
# library("plotrix")
library("R.devices")
```

Import the data files 
```{r}
# set working directory to RAnalysis

setwd("/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis")
getwd()

#treatment information
treatmentinfo <- read.csv("/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Data/Pacuta/Pacuta_RNA_seq_metadata.csv", header = TRUE, sep = ",") # pacuta
treatmentinfo <- read.csv("/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Data/Mcap/Mcap_RNAseq_metadata.csv", header = TRUE, sep = ",") # mcap
rownames(treatmentinfo) <- treatmentinfo$sample_num
str(treatmentinfo)
head(treatmentinfo)

#gene count matrix
gcount <- as.data.frame(read.csv("../../stringtie_assembly/counts_matrices/Pacuta-v2/gene_count_matrix2.csv", row.names="gene_id")) # pacuta
gcount <- as.data.frame(read.csv("/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/stringtie_assembly/counts_matrices/Mcap/redo/gene_count_matrix2.csv", row.names="gene_id")) # mcap

gcount<- tibble::rownames_to_column(gcount, "gene_id")
dim(gcount)
head(gcount)
colSums(gcount)
colnames(gcount)

```

##### Pre-filter gene counts
## Pre-filtering our dataset to reduce the memory size dataframe, increase the speed of the transformation and testing functions, and improve quality of statistical analysis by removing low-coverage counts. Removed counts could represent outliers in the data and removing these improves sensitivity of statistical tests. We will filter out low coverage samples. 

```{r}
#Do we want to do any filtering?

#why do we want to filter?

#what do we want to filter?

#how do we want to filter

#Prefiltering versus DESeq2 default filtering
#http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#indfilt


filt <- filterfun(pOverA(0.5,10))

#create filter for the counts data
gfilt <- genefilter(gcount, filt)


#identify genes to keep by count filter
gkeep <- gcount[gfilt,]


#identify gene lists
gn.keep <- rownames(gkeep)


#gene count data filtered in PoverA, P percent of the samples have counts over A
gcount_filt <- as.data.frame(gcount[which(rownames(gcount) %in% gn.keep),])
head(gcount_filt)
dim(gcount_filt) 

# save gcount filt as a .csv file for GOSeq
write.csv(gcount_filt, '/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919_Pacuta2_filtered_gene_count.csv')

# check maximum values for each sample and record corresponding row name 

max(gcount_filt)
max(gcount_filt$X1041.2)
max(gcount_filt$X1471.2)
max(gcount_filt$X1637.2)
max(gcount_filt$Sample1.2)
max(gcount_filt$Sample2.2)
max(gcount_filt$Sample3.2)
gcount_filt[which(gcount_filt$X1041.2 == 218587 | gcount_filt$X1471.2 == 244867 | gcount_filt$X1637.2 == 203630 | gcount_filt$Sample1.2 == 55184), ]

gcount_filt <- subset(gcount_filt, X1041.2 != 218587)

max(gcount_filt_no_max)
# MSTRG.4996
data[ which( data$V1 > 2 | data$V2 < 4) , ]
max(gcount_filt$X1041.2)
gcount_filt[which(gcount_filt$X1041.2 == 218587), ]
```

## create a scatterplot to visualize gene counts 

```{r}
# plot gene abundances for filted gene counts
gcount_filt <- cbind(rownames(gcount_filt), data.frame(gcount_filt, row.names=NULL))
colnames(gcount_filt) <- c("gene_id", "1041", "1548", "1628", "Sample4", "Sample5", "Sample6") # mcap DOUBLE CHECK
colnames(gcount_filt) <- c("gene_id", "1041", "1471", "1637", "Sample1", "Sample2", "Sample3") # pacuta 

gcount_filt_long <- gcount_filt %>%
  pivot_longer(2:7,
               names_to = "sample", values_to = "gene_count") # by using a pivot function we can make our data "long format" so it is easier to plot 
gcount_filt_long_plot <- ggplot(gcount_filt_long, aes(x=gene_id, y=gene_count,color=sample)) +
  geom_point() +
  labs(x = "gene id", y = "gene abundance")+
  scale_color_manual(values=c("pink","pink","pink","orange","orange","orange"))

print(gcount_filt_long_plot)

# ggsave("Pacuta-gcount-filt-histogram.pdf") not a figure we will keep for ms
## note: isolate high abundance transcripts and find out what they are 
```
## this is optional -- may delete for final script 

```{r}
# remove rows with zeros for all values 

gcount.no.zeros <- gcount_filt %>% filter(X1041.2 > 0, X1471.2 > 0, X1637.2 > 0, Sample1.2 > 0, Sample2.2 > 0, Sample3.2 > 0 )
gcount.no.zeros <- gcount_filt %>% filter(X1101.2 > 0, X1548.2 > 0, X1628.2 > 0, Sample4.2 > 0, Sample5.2 > 0, Sample6.2 > 0 )

# subset for rows that have all zeros for one kit and not the other 

gcount.polya.zero <- gcount_filt %>% filter("1101" == 0)
polya_zero_long <- gcount.polya.zero %>%
  pivot_longer(2:7,
               names_to = "sample", values_to = "gene_count")
view(polya_zero_long)

high_abundance_long_plot <- ggplot(abundance_long, aes(x=gene_id, y=gene_count,color=sample)) +
  geom_point() + 
  ggtitle("*Pocillopora acuta* highly abundant gene counts")+
  labs(x = "gene id", y = "gene abundance")+
  scale_color_manual(values=c("pink","pink","pink","orange","orange","orange"))
print(high_abundance_long_plot)
```
## this is optional -- may delete for final script 

```{r}
gcount.ribofree <- gcount_filt %>% filter(Sample4.2.gtf == 0, Sample5.2.gtf == 0, Sample6.2.gtf == 0 )

# subset for highly abundant transcripts

gcount.high.abundance <- gcount_filt %>% filter(1101 >= 336, 1548 >= 306, 1628 >= 264, Sample4 >= 30, Sample5 >= 27, Sample6 >= 29)
gcount.high.abundance <- cbind(rownames(gcount.high.abundance), data.frame(gcount.high.abundance, row.names=NULL))
colnames(gcount.high.abundance) <- c("gene_id", "Sample1101", "1548", "1628", "Sample4", "Sample5", "Sample6")

abundance_long <- gcount.high.abundance %>%
  pivot_longer(2:7,
               names_to = "sample", values_to = "gene_count")
view(abundance_long)

high_abundance_long_plot <- ggplot(abundance_long, aes(x=gene_id, y=gene_count,color=sample)) +
  geom_point() + 
  ggtitle("*Pocillopora acuta* highly abundant gene counts")+
  labs(x = "gene id", y = "gene abundance")+
  scale_color_manual(values=c("pink","pink","pink","orange","orange","orange"))
print(high_abundance_long_plot)

ggsave("Pacuta-high-abundance-histogram.pdf")
```

# basic heatmap of correlation between all possible columns 
```{r}
library(ggcorrplot)
library(Hmisc)


corr <- cor(gcount_filt)
print(corr)
ggcorrplot(corr, hc.order = TRUE, outline.color = "white")
ggsave("20220919_Pacuta_filtered_genecount_corr_plot.pdf")
```

# optional heatmap -- might delete
```{r}
high_abundance_corr_plot <- ggcorrplot(cor(gcount.high.abundance[2:7]))+
  ggtitle("Pocillopora acuta highly abundant counts correlation by sample")
ggsave("20220919_Pacuta_high_abundance_corr_plot.pdf")
print(high_abundance_corr_plot)

```

# create a pairs panel for P over A counts
```{r}
library(psych)
library(graphics)
# creating pairs panels 

# unfiltered from counts data 

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919acuta_PoverApairs.pdf")
pairs.panels(gcount_filt[1:6], hist.col = "pink", cex.labels = 1.25)
dev.off()
```

# options may delete 

```{r}
# zeros in both data sets removed 
gcount.no.zeros <- filter(gcount, X1041.gtf > "0", X1471.gtf > "0", X1637.gtf  > "0", Sample1.gtf > "0", Sample2.gtf > "0", Sample3.gtf > "0" )# use for pacuta

gcount.no.zeros <- filter(gcount, X1101 > "0", X1548 > "0", X1628  > "0", Sample4 > "0", Sample5 > "0", Sample6 > "0" ) # use for mcap

## now create pairs panels

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220610_Pacuta_nozeros_pairs.pdf")
no.zeros.pairs <- pairs.panels(gcount.no.zeros, hist.col = "pink", cex.labels = 1.25)
dev.off()

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Mcap/20220610_Mcap_nozeros_pairs.pdf")
no.zeros.pairs <- pairs.panels(gcount.no.zeros, hist.col = "pink", cex.labels = 1.25)
dev.off()


# pairs panel using filtered for p over a (0.5,10)

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220610_Pacuta_filtered_pairs.pdf")
gcount.filt.pairs <- pairs.panels(gcount_filt, hist.col = "orange", cex.labels = 1.25)
dev.off()

```

 Create a DESeqDataSet design from gene count matrix and labels. Here we set the design to look at treatment to test for any differences in gene expression attributed to library prep method.
 not using DESeq due to random factors, but can still use this to visualize clusters

```{r}
# Set DESeq2 design

gdds <- DESeqDataSetFromMatrix(countData = gcount_filt,
                              colData = treatmentinfo,
                              design = ~method)
```

#### Visualize gene count data

We're looking to see if the samples of the same treatments cluster

##### Log-transform the count data
First we are going to log-transform the data using a variance stabilizing transforamtion (vst). This is only for visualization purposes. Essentially, this is roughly similar to putting the data on the log2 scale. It will deal with the sampling variability of low counts by calculating within-group variability (if blind=FALSE). Importantly, it does not use the design to remove variation in the data, and so can be used to examine if there may be any variability do to technical factors such as extraction batch effects.

To do this we first need to calculate the size factors of our samples. This is a rough estimate of how many reads each sample contains compared to the others. In order to use VST (the faster log2 transforming process) to log-transform our data, the size factors need to be less than 4. Otherwise, there could be artifacts in our results.
```{r}
SF.gdds <- estimateSizeFactors(gdds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than four to use vst
print(sizeFactors(SF.gdds)) #View size factors
```

Our size factors are all less than 4, so we can use VST!
```{r}
gvst <- vst(gdds, blind=FALSE) #apply a variance stabilizing transformation to minimize effects of small counts and normalize library size
```

#### Principal component plot of samples
```{r}
gPCAdata <- plotPCA(gvst, intgroup = c("method"), returnData=TRUE)
percentVar <- round(100*attr(gPCAdata, "percentVar")) #plot PCA of samples with all data
pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919-Pacuta2-PCAplot.pdf")
PCAplot <- ggplot(gPCAdata, aes(PC1, PC2, color=method)) + 
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  #scale_color_manual(values = c(PolyA="slateblue", RiboFree="indianred3")) +
  coord_fixed() +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     #panel.grid.major = element_blank(), #Set major gridlines
                     #panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()) + #Set the plot background
  theme(legend.position = ("none")) + #set title attributes
  ggtitle("Pocillopora acuta")

#create file to save locally in the output folder
PCAplot
dev.off()

## sometimes dev.off() doesn't work, you can also use ggsave
ggsave(path = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta", filename = "PacutaPCAplot.pdf")


```

##### Run DE analysis for overall example 

#Run differential expression test using a Wald model. 
```{r, message = FALSE}
DEGSeq2_wald <- DESeq(gdds, test = "Wald") #run differential expression test by group using the Wald model
```

#Explore significant p-values for treatment
```{r, message = FALSE}
DEGSeq2.results <- results(DEGSeq2_wald, contrast=c("method","PolyA","RiboFree")) 
head(DEGSeq2.results)
sum(DEGSeq2.results$padj < 0.05, na.rm=TRUE) #How many adjusted p-values were less than 0.05
# Pacuta 6479

sum(DEGSeq2.results$padj < 0.05 & DEGSeq2.results$log2FoldChange > 0, na.rm=TRUE) # how many adjusted p-values were more than 0.05 

sum(DEGSeq2.results$padj < 0.05 & DEGSeq2.results$log2FoldChange < 0, na.rm=TRUE) # how many adjusted p-values were less than 0.05


#create data frame with only DESeq2 DEG's with < 0.05 significance 

DEGSeq2.sig.results <- subset(DEGSeq2.results, padj < 0.05)
DEGSeq2.sig.results <- as.data.frame(DEGSeq2.sig.results)
DEGSeq2.sig.results$gene_id <-rownames(DEGSeq2.sig.results)
write.csv(DEGSeq2.sig.results, '/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919_Pacuta-DEGSeq2.sig.results.csv')  

```

## Visualize differentially expressed genes from DESeq2
```{r}
# Subset list of genes by those which padj>0.05
DEGSeq2.output <- as.data.frame(DEGSeq2.sig.results)
dim(DEGSeq2.output)
DEGSeq2.results.all <- DEGSeq2.sig.results
DEGSeq2 <- rownames(DEGSeq2.results.all) # list all gene names 
DEGSeq2 <- unique(DEGSeq2) # select only unique gene names 
DEGSeq2_list <- gdds[which(rownames(gdds) %in% DEGSeq2)] # filter gdds DESeq2 object by unique gene names
dim(DEGSeq2_list) # 15 x 15
view(DEGSeq2_list)
print(counts(DEGSeq2_list))

# As determined above, size factors all less than 4, so proceed with VST
# apply a variance stabilizing transformation to minimize effects of small counts and normalize library size

DEGvst <- vst(DEGSeq2_list, blind = FALSE) #nsub = nrow(counts(DEGSeq2_list)))

dim(DEGvst) # 15 x 15
print(assay(DEGvst)) # look at vst-transformed gene count data 

# Plot heat map with diff expressed genes
# Make a matrix for computing similarity
mat <- assay(DEGvst)#[DEG.results.all, ] # make an expression object
mat <- mat - rowMeans(mat) #difference in expression compared to average across all samples
dim(mat)
ann_colors <- list(treatment= c(PolyA="black", RiboFree="cadetblue3"))
df_DEGSeq2 <- as.data.frame(colData(DEGvst)[c("method")]) #make dataframe for column naming and associated treatment
DEGSeq2_heatmap <- pheatmap(mat, scale= "row", legend=TRUE, annotation_legend=TRUE, annotation_col=df_DEGSeq2, annotation_colors = ann_colors,
                            clustering_distance_rows="euclidean", clustering_method = "average",
                            show_rownames =FALSE,
                            show_colnames =TRUE,
                            cluster_cols = TRUE)

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919-Pacuta2_heatmap.pdf")
DEGSeq2_heatmap
dev.off()

```


```{r}
# PCA plot of diff-expressed genes 
DEGSeq2_PCAdata <- plotPCA(DEGvst, intgroup = c("method"), returnData=TRUE)
percentVar_pca <- round(100*attr(DEGSeq2_PCAdata, "percentVar")) #plot PCA of samples with all data
DEGSeq2_PCA_plot <- ggplot(DEGSeq2_PCAdata, aes(PC1, PC2, color=method)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar_pca[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_pca[2],"% variance")) +
  scale_color_manual(values = c(PolyA="black", RiboFree="cadetblue3")) +
  coord_fixed() +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
        #panel.grid.major = element_blank(), #Set major gridlines
        #panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank())+ #Set the plot background
  ggtitle("Pocillopora acuta DEG")

DEGSeq2_PCA_plot

ggsave(path = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta", filename = "Pacuta_DEGSeq2_PCA_plot.pdf")

# PCA plot is of deferentially expressed genes only

# Save results

write.csv(counts(DEGSeq2_list), "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220828_Pacuta-v2DEGSeq2_list_unique.csv")



```
```{r}
# subset gcount_filt for DEGs from DEG dataframe to create the basic heatmap and the pairs panel
rownames(gcount_filt) # check that rownames have been retained from original loading of df
DEGSeq_list <- rownames(DEGSeq2.sig.results) # create a vector from the rownames from the significant results
DEG <- as.data.frame(gcount_filt[which(rownames(gcount_filt) %in% DEGSeq_list),])

```

# repeat basic heatmap for DEGs
```{r}
library(ggcorrplot)
library(Hmisc)


corr_plot <- ggcorrplot(cor(DEG[1:6])) 
print(corr_plot)
ggsave("20220919_Pacuta_DEG_corr_plot.pdf")
```


# repeat pairs panel for DEGs
```{r}
library(psych)
library(graphics)
# creating pairs panels 

# unfiltered from counts data 

pdf(file = "/Users/laurenzane/Desktop/Putnam_Lab/Projects/Express_Compare/RAnalysis/Output/Pacuta/20220919_Pacuta_DEGpairs.pdf")
pairs.panels(DEG[1:6], hist.col = "pink", cex.labels = 1.25)
dev.off()
```



# GO Seq preparing and loading relevant dataframes
```{r}
library(goseq) #installing package
## recall that gcount_filt is the data frame contains all genes that are expressed (pOverA(0.5,10)
## DEGSeq2.sig.results is the data frame that contains all differentially expressed genes 

head(gcount_filt)
gcount_filt2 <- tibble::rownames_to_column(gcount_filt, "gene_id") # creates column for gene id in the filtered gene count data frame
head(gcount_filt2)
# gcount_filt2$gene_id <- gsub("t1","", as.character(gcount_filt2$gene_id))# removes t1 from gene id
# gcount_filt2$gene_id <- gsub("t2","", as.character(gcount_filt2$gene_id))# removes t2 from gene id
# gcount_filt2$gene_id <- gsub("\\.*$","", as.character(gcount_filt2$gene_id)) # removes . from end of gene id
print(gcount_filt2$gene_id)

head(DEGSeq2.sig.results)
DEGSeq2.sig.results2 <- tibble::rownames_to_column(DEGSeq2.sig.results, "gene_id")# create column for gene id in DEG significant results dataframe 
# seems to already have gene_id
print(colnames(DEGSeq2.sig.results))
head(DEGSeq2.sig.results)
# head(DEGSeq2.sig.results2)
# DEGSeq2.sig.results2$gene_id<-gsub("t1","",as.character(DEGSeq2.sig.results2$gene_id))#removes t1 from gene id
# DEGSeq2.sig.results2$gene_id<-gsub("t2","",as.character(DEGSeq2.sig.results2$gene_id))# removes t2 from gene id
# DEGSeq2.sig.results2$gene_id<-gsub("\\.*$","",as.character(DEGSeq2.sig.results2$gene_id))#removes period at end of gene id 
print(DEGSeq2.sig.results2$gene_id)


```





